Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-24 paper/paper.tex
pass through section 4

\section{Semantic Typing and Consistency}
\label{sec:logrelproof}
\yl{todo: rename all variables from i,j to n}
\begin{figure}[h]
\[
\begin{array}{lcl}

      [[rho |= G]] &:= & \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
                     } [[InterpR j (G i < up Suc i > ) { rho } S ]] \text{, then } [[rho i in S]] \\
      [[G |= a : A]] &:= & \forall [[rho]], \text{ if }[[rho |=
                       G]]\text{ then there exists some } [[j]] \text{
                       and } [[S]] \text{ such that } [[InterpR j A
                       {rho} S]] \text{ and } [[a {rho} in S]] \\
      [[|= G]] &:= & \forall [[i < |G|]], \exists [[j]], [[drop Suc i G |= G i : Set j]]
\end{array}
\]
  \caption{Semantic Typing for \lang}
  \label{fig:semtyping}
\end{figure}

@@

% \scw{What is the motivation for this section? Why do we want to generalize
% the logical relation?}
In this section, we define the notion of semantic typing based on the
logical relation we have defined in Section~\ref{sec:logreldep} and
prove the fundamental lemma, which states that syntactic typing
implies semantic typing.

@@

% To generalize our logical relation to open
% terms, we define the semantic typing judgment by closing the open
% terms with a substitution whose codomain consists of terms that
% respect the interpretation of the types from the context.
The full
definitions of well-formed substitution ($[[rho |= G]]$), semantic
typing ($[[ G |= a : A]]$), and semantic context well-formedness
($[[|= G]]$) are presented in Figure~\ref{fig:semtyping}.

The $[[rho |= G]]$ notation denotes the semantic well-formedness of
a substitution $[[rho]]$ with respect to a context
$[[G]]$. Intuitively, for variable $[[i]]$ in the context (thus the
restriction $[[i < | G |]]$), $[[rho i]]$ is a term that inhabits all
possible interpretations of the type $[[G i]]$ from the
context.
Since we are in a dependently typed system, there
may be variables in $[[G i]]$, too. The type $[[(G i < up Suc i > ) {
  rho }]]$ applies $[[rho]]$ to replace those variables. The shifting
operation $[[up Suc i]]$ can be viewed as an artifact of the de Bruijn
representation to lift the type $[[G i]]$ to the right scope before
substitution can be carried out.
Finally, despite the for all quantification over the interpreted set
$[[S]]$ and the universe level $[[j]]$ in the definition of $[[rho |=
G]]$, we only need show that there exists some $[[j]]$ and $[[S]]$
such that $[[InterpR j (G i < up Suc i > ) { rho } S]]$ and $[[rho i
in S]]$. This idea is encapsulated in the second of the two structural
law about $[[rho |= G]]$.
\begin{lemma}[Well-formed $[[rho]]$ empty\footnote{\dotv{soundness.v}{$\gamma$\_ok\_nil}}]
  \label{lemma:rhowfempty}
  $[[rho |= G]]$ whenever $[[G]]$ is the empty context.
\end{lemma}
\begin{lemma}[Well-formed $[[rho]]$ cons\footnote{\dotv{soundness.v}{$\gamma$\_ok\_cons}}]
  If $[[InterpR i A S]]$, $[[a in S]]$, and $[[rho |= G]]$, then
  $[[rho .: a |= G ++ A]]$.
\end{lemma}

We refer to $[[G |= a : A]]$ as semantic well-typedness. Its definition says the term $[[a]]$ can be
semantically typed $[[A]]$ under the context $[[G]]$ if for all
substitutions $[[rho]]$ such that $[[rho |= G]]$, the type $[[A { rho }]]$ can be interpreted
as the set $[[S]]$, and $[[a { rho } in S]]$. Our definition of
semantic well-typedness is standard, though dependent types add a
small twist that we apply the $[[rho]]$ to $[[A]]$ and require that
$[[A { rho }]]$ has some interpretation, which is not always the case
for arbitrary types.

Analogous to $[[|- G]]$, the syntactic well-formedness of the typing
context, $[[|= G]]$ defines when a context is semantically
well-formed. In the
definition of $[[|= G]]$, we use the notation $[[drop i G]]$ to denote
the typing context obtained by dropping the last $[[i]]$ elements of
$[[G]]$. When $[[G]]$ has less than $[[i]]$ elements, $[[drop i G]]$
returns the empty list. Intuitively, the definition of $[[|= G]]$
simply says all types in the context must correspond to some set
according to the logical relation. The truncation to the context
$[[G]]$ is due to the different scoping of the types in the
context. An alternative way of defining $[[|= G]]$ would look more
similar to $[[|- G]]$, though we find it easier to define $[[|= G]]$
through for all quantification of $[[i]]$ and recover the structural
rules as lemmas.
\begin{lemma}[Semantic context well-formedness cons]
  \label{lemma:semwffcons}
  If $[[|= G]]$ and $[[G |= A : Set i]]$, then $[[|= G ++ A]]$.
\end{lemma}
\begin{lemma}[Semantic context well-formedness empty]
  \label{lemma:semwffempty}
  $[[|= G]]$ holds when $[[G]]$ is empty.
\end{lemma}

@@

Next, we show some non-trivial cases of the fundamental theorem as
top-level lemmas.
\begin{lemma}[ST-Var]
  \label{lemma:stvar}
  If $[[|= G]]$ and $[[i < |G|]]$, then $[[G |= var i : G i < up Suc i  >]]$.
\end{lemma}
\begin{proof}
  Suppose $[[rho |= G]]$. By the definition of semantic typing, we
need to show that there exists some $[[j]]$ and $[[S]]$ such that
  \begin{itemize}
  \item $[[InterpR j G i < up Suc i > { rho } S]]$
  \item $[[rho i in S]]$
  \end{itemize}
  From the definition of $[[|= G]]$, we know that there exists some
  $[[j]]$ such that $[[drop Suc i G |= G i : Set j]]$. This is not
  immediately useful since $[[rho |= G]]$, not the truncated
  context. We claim, without giving the proof, that from $[[drop Suc i
  G |= G i  : Set j]]$, we can derive $[[G |= G i {up Suc i} : Set
  j]]$. A rigorous justification would require us to formally state
  and show renaming/weakening lemma for semantic typing. We instead
  provide an informal justification since the proof of renaming
  is uninteresting and only requires unfolding definitions, but its
  statement introduces unnecessary line noise related to shifting.

  Intuitively, a substitution $[[rho]]$
  that closes over a context should always induce a truncated
  substituion for each truncated context. As a result, if
  some term $[[b]]$ is semantically well-typed under some truncated
  context $[[G]]$, then it must also be semantically well-typed under
  the context $[[G]]$ up to shifting.

  Knowing that $[[G |= G i {up Suc i} : Set j]]$ holds, we can then
  apply Lemma~\ref{lemma:setinv} to show that
  $[[InterpR j G i < up Suc i > { rho } S]]$. It now
  suffices to show $[[rho i in S]]$, but that is immediate from the
  definition of $[[rho |= G]]$.
\end{proof}

@@

\begin{theorem}[The Fundamental Theorem\footnote{\dotv{soundness.v}{soundness}}]
  \label{theorem:soundness}\leavevmode
  \begin{itemize}
  \item If $[[G |- a : A]]$, then $[[G |= a : A]]$.
  \item If $[[|- G]]$, then $[[|= G]]$.
  \end{itemize}
\end{theorem}
\begin{proof}
  Proof by mutual induction over the derivation of $[[G |- a :
  A]]$ and $[[|- G]]$.   The cases related to context well-formedness immediately follows
  from Lemma~\ref{lemma:semwffcons}.
Lemmas~\ref{lemma:stvar},~\ref{lemma:stset},~\ref{lemma:stpi},~\ref{lemma:stabs},~\ref{lemma:stapp}
  can be used to discharge their syntactic counterpart
  (e.g. Lemma~\ref{lemma:stabs} for case \rref{T-Abs}). The remaining
  cases not covered by the lemmas are similar to the ones already
  shown or simpler and therefore omitted from the text.
\end{proof}

@@

\begin{proof}
  Suppose $[[empty |- a : Void]]$ is derivable, then by the
  fundamental lemma, we have $[[empty |= a : Void]]$, which states
  that for all $[[rho |= empty]]$, and for all $[[j]]$, $[[S]]$ such
  that $[[InterpR j Void S]]$, we have $[[a {rho} in S]]$. By
  Lemma~\ref{lemma:rhowfempty}, any $[[rho]]$ we pick trivially
  satisfies $[[rho |= G]]$. For convenience, we pick $[[rho]]$ as
  $[[idtm]]$ (though any $[[rho]]$ would work since
  $[[empty |- a : Void]]$ ensures there is no free variable in
  $[[a]]$), we have $[[a {idtm}]] = [[a in S]]$. By the $[[Void]]$
  case of the inversion property (Lemma~\ref{lemma:interpinv}), we
  know that $[[S]]$ must be the empty set and therefore contradicts
  the assupmtion that $[[a in S]]$. Done.
\end{proof}

>>>>

\section{Semantic Typing and Consistency}
\label{sec:logrelproof}
\yl{todo: rename all variables from i,j to n}
\scw{Would it make sense to define the notation $a \in [\![A]\!]^i$ when
there exists some $S$ such that $[[InterpR i A S]]$ and $[[a in S]]$ ?}
% \begin{figure}[h]
% \[
% \begin{array}{lcl}

%       [[rho |= G]] &:= & \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
%                      } [[InterpR j (G i < up Suc i > ) { rho } S ]] \text{, then } [[rho i in S]] \\
%       [[G |= a : A]] &:= & \forall [[rho]], \text{ if }[[rho |= G]]
%    \text{ then there exists some } [[j]] \text{
%                        and } [[S]] \text{ such that } [[InterpR j A {rho} S]]
%        \text{ and } [[a {rho} in S]] \\
%       [[|= G]] &:= & \forall [[i < |G|]], \exists [[j]], [[drop Suc i G |= G i : Set j]]
% \end{array}
% \]
%   \caption{Semantic Typing for \lang}
%   \label{fig:semtyping}
% \end{figure}x

@@

In this section, we show that all closed, well-typed terms are contained
within their type-indexed sets. In other words, $[[empty |- a : A]]$ implies
$[[InterpR i A S]]$ and $[[a in S]]$.  This result gives us consistency
because we know that $[[InterpR i Void S]]$ is defined, and that $[[S]]$ must
be the empty set. Therefore, if there were some closed, well-typed term of type
$[[Void]]$, it would need to be a member of the empty set, a contradiction.

To prove this result, we define a notion of semantic typing based on the
logical relation we have defined in Section~\ref{sec:logreldep} and prove the
fundamental lemma, which states that syntactic typing implies semantic typing.
Semantic typing extends our logical relation from being a (type-indexed)
family of predicates on closed terms, to a type-indexed family of predicates
on open terms.

@@

\begin{definition}[Semantic well-formed substitution\footnote{\dotv{soundness.v}{$\rho$\_ok}}]
Define $[[rho |= G]]$ when 
\[ \forall i\ j\ S, \text{ if }[[i < |G|]]\text{ and
                     } [[InterpR j (G i < up Suc i > ) { rho } S ]] \text{, then } [[rho i in S]] \]
\end{definition}

% To generalize our logical relation to open
% terms, we define the semantic typing judgment by closing the open
% terms with a substitution whose codomain consists of terms that
% respect the interpretation of the types from the context.
% The full
% definitions of well-formed substitution ($[[rho |= G]]$), semantic
% typing ($[[ G |= a : A]]$), and semantic context well-formedness
% ($[[|= G]]$) are presented in Figure~\ref{fig:semtyping}.

The $[[rho |= G]]$ notation denotes the semantic well-formedness of a
substitution $[[rho]]$ with respect to a context $[[G]]$. Intuitively, for any
variable $[[i]]$ in the context (thus the restriction $[[i < | G |]]$),
$[[rho i]]$ is a term that inhabits all possible interpretations of the type
$[[G i]]$ from the context.

Since we are in a dependently typed system, there
may be variables in $[[G i]]$, too. The type $[[(G i < up Suc i > ) {
  rho }]]$ applies $[[rho]]$ to replace those variables. The shifting
operation $[[up Suc i]]$ can be viewed as an artifact of the de Bruijn
representation to lift the type $[[G i]]$ to the right scope before
substitution can be carried out.

Finally, despite the for all quantification over the interpreted set
$[[S]]$ and the universe level $[[j]]$ in the definition of $[[rho |=
G]]$, we only need show that there exists some $[[j]]$ and $[[S]]$
such that $[[InterpR j (G i < up Suc i > ) { rho } S]]$ and $[[rho i
in S]]$. This idea is encapsulated in the second of the two structural
law about $[[rho |= G]]$.
\begin{lemma}[Well-formed $[[rho]]$ empty\footnote{\dotv{soundness.v}{$\rho$\_ok\_nil}}]
  \label{lemma:rhowfempty}
  $[[rho |= G]]$ whenever $[[G]]$ is the empty context.
\end{lemma}
\begin{lemma}[Well-formed $[[rho]]$ cons\footnote{\dotv{soundness.v}{$\rho$\_ok\_cons}}]
  If $[[InterpR i A S]]$, $[[a in S]]$, and $[[rho |= G]]$, then
  $[[rho .: a |= G ++ A]]$.
\end{lemma}

We next define semantic well-typedness.

\begin{definition}[Semantic typing\footnote{\dotv{soundness.v}{SemWt}}]
Define $[[G |= a : A]]$  when
\[ \forall [[rho]], \text{ if }[[rho |=
                       G]]\text{ then there exists some } [[j]] \text{
                       and } [[S]] \text{ such that } [[InterpR j A
                       {rho} S]] \text{ and } [[a {rho} in S]]  \]
\end{definition}

This definition says the term $[[a]]$ can be semantically typed $[[A]]$ under
the context $[[G]]$ if for all substitutions $[[rho]]$ such that
$[[rho |= G]]$, the type $[[A { rho }]]$ can be interpreted as the set
$[[S]]$, and $[[a { rho } in S]]$. Our definition of semantic well-typedness
is standard, though dependent types add a small twist that we apply the
$[[rho]]$ to $[[A]]$ and require that $[[A { rho }]]$ has some interpretation.

Finally, we define semantic well-formedness for contexts, analogous to the 
relation $[[|-G]]$.

\begin{definition}[Semantic context well-formedness\footnote{\dotv{soundness.v}{SemWff}}]
Define $[[|= G]]$ when
\[
   \forall [[i < |G|]],\text{ there exists some } [[j]] \text{ such that } [[drop Suc i G |= G i : Set j]] \]
\end{definition}

In this definition, we use the notation $[[drop i G]]$ to denote
the typing context obtained by dropping the last $[[i]]$ elements of
$[[G]]$. When $[[G]]$ has fewer than $[[i]]$ elements, $[[drop i G]]$ returns
the empty list. Intuitively, $[[|= G]]$ says that all
types in the context must correspond to some set according to the logical
relation. The truncation to the context $[[G]]$ is due to the telescopic
scoping of the types in the context. An alternative approach would be to define $[[|= G]]$
inductively, similar to $[[|- G]]$, but we find it easier to use this definition and
recover the structural rules as lemmas.
\begin{lemma}[Semantic context well-formedness cons]
  \label{lemma:semwffcons}
  If $[[|= G]]$ and $[[G |= A : Set i]]$, then $[[|= G ++ A]]$.
\end{lemma}
\begin{lemma}[Semantic context well-formedness empty]
  \label{lemma:semwffempty}
  $[[|= G]]$ holds when $[[G]]$ is empty.
\end{lemma}

@@

Next, we show some non-trivial cases of the fundamental theorem as
top-level lemmas. For example, we can define the semantic analogue to the 
syntactic typing rule for variables (\rref{T-Var}).
\begin{lemma}[ST-Var]
  \label{lemma:stvar}
  If $[[|= G]]$ and $[[i < |G|]]$, then $[[G |= var i : G i < up Suc i  >]]$.
\end{lemma}
\begin{proof}
  Suppose $[[rho |= G]]$. By the definition of semantic typing, we
need to show that there exists some $[[j]]$ and $[[S]]$ such that
  \begin{itemize}
  \item $[[InterpR j G i < up Suc i > { rho } S]]$
  \item $[[rho i in S]]$
  \end{itemize}
  From the definition of $[[|= G]]$, we know that there exists some
  $[[j]]$ such that $[[drop Suc i G |= G i : Set j]]$. This is not
  immediately useful since $[[rho |= G]]$, not the truncated
  context. We claim, without giving the proof, that from $[[drop Suc i
  G |= G i  : Set j]]$, we can derive $[[G |= G i {up Suc i} : Set
  j]]$. A rigorous justification would require us to formally state
  and show renaming/weakening lemma for semantic typing. We instead
  provide an informal justification since the proof of renaming
  is uninteresting and only requires unfolding definitions, but its
  statement introduces unnecessary line noise related to shifting.

  Intuitively, a substitution $[[rho]]$
  that closes over a context should always induce a truncated
  substituion for each truncated context. As a result, if
  some term $[[b]]$ is semantically well-typed under some truncated
  context $[[G]]$, then it must also be semantically well-typed under
  the context $[[G]]$ up to shifting.
%
  Knowing that $[[G |= G i {up Suc i} : Set j]]$ holds, we can then
  apply Lemma~\ref{lemma:setinv} to show that
  $[[InterpR j G i < up Suc i > { rho } S]]$. It now
  suffices to show $[[rho i in S]]$, but that is immediate from the
  definition of $[[rho |= G]]$.
\end{proof}

@@

\begin{theorem}[The Fundamental Theorem\footnote{\dotv{soundness.v}{soundness}}]
  \label{theorem:soundness}\leavevmode
  \begin{itemize}
  \item If $[[G |- a : A]]$, then $[[G |= a : A]]$.
  \item If $[[|- G]]$, then $[[|= G]]$.
  \end{itemize}
\end{theorem}
\begin{proof}
  Proof by mutual induction over the derivation of $[[G |- a :
  A]]$ and $[[|- G]]$.   The cases related to context well-formedness immediately follows
  from Lemma~\ref{lemma:semwffcons}. The semantic typing rules
(Lemmas~\ref{lemma:stvar},~\ref{lemma:stset},~\ref{lemma:stpi},~\ref{lemma:stabs},~\ref{lemma:stapp})
  can be used to discharge their syntactic counterparts
  (e.g. Lemma~\ref{lemma:stabs} for case \rref{T-Abs}). The remaining
  cases not covered by the lemmas are similar to the ones already
  shown.
\end{proof}

@@

\begin{proof}
  Suppose $[[empty |- a : Void]]$ is derivable, then by the
  fundamental lemma, we have $[[empty |= a : Void]]$, which states
  that for all $[[rho |= empty]]$, and for all $[[j]]$, $[[S]]$ such
  that $[[InterpR j Void S]]$, we have $[[a {rho} in S]]$. By
  Lemma~\ref{lemma:rhowfempty}, any $[[rho]]$ we pick trivially
  satisfies $[[rho |= G]]$. For convenience, we pick $[[rho]]$ as
  $[[idtm]]$, though any $[[rho]]$ would work since
  $[[empty |- a : Void]]$ ensures there is no free variable in
  $[[a]]$. We have $[[a {idtm}]] = [[a in S]]$. By the $[[Void]]$
  case of the inversion property (Lemma~\ref{lemma:interpinv}), we
  know that $[[S]]$ must be the empty set and therefore contradicts
  the assumption that $[[a in S]]$.
\end{proof}
